---
title: "Rethinking | Take 3"
author: "Lukas Feuer"
output:
  html_notebook:
    theme: cosmo
    highlight: pygments
    toc: yes
    toc_float: yes
---

This is my third attempt at completing the full course. I am now waching the excellent companion lectures by Richard Mecelreath on [YouTube](https://youtu.be/guTdrfycW2Q?t=1301). My plan is to code along, starting with lecture 2 and take notes (most likely when I am about to go bejond my second attempt in the book)

**The Plan** is to finally get through the full course, read the book later if needed

# 02 | Globe Tossing

* No point estimates
* Intervalls are only for brief reporting, not for error control 

```{r Globe tossing example}
# Globe tossing example | Grid Approximation

p_grid = seq(0, 1, len = 1000)

prob_p <- rep(1, 1000)

prob_data <- dbinom(6, 9, prob = p_grid)

poster  <- prob_data * prob_p
poster <- poster/sum(poster)
```

* Samples from posterior to summarize 
* Creating the Posterior Predictive --> explanation and visual around 1:00:00 (2022, Lecture 02) 
* Posterior Predicitve (whole distribution) first goal to report 

```{r Sample from the Posterior}
samples <- sample(p_grid, prob = poster, size = 1e4, replace = TRUE)

#library(rethinking)
#plot(samples)
#dens(samples)

w <- rbinom(1e4, 9, prob = samples)

# POSTERIOR Predictive (normaization missing?)
hist(w)
```

* **PRIOR Predicitve** works the same but with samples from the prior (before seeing the data)
* Very useful in complex models if we have to choose priors and don't understand them in isolation 
* Makes it understandable, what Priors "mean" --> choose them knowingly, not by default 


# 03 | Geocentric Models



```{r}

```

